Supplementary Material for Asymmetry paper

### 2020-07-27 Supplementary/Sup_mat.Rmd
corrected spelling mistakes

These supplementary materials present full details of all R packages used in the analysis; a detailed explaination of our simulation-based power analysis; and Bayesian models of the results discussed in the main manuscript. In addition, we also present details a further, related experiment using eye-tracking and a probability matching paradigm.

@@

The fit with the empirical data is reasonable, although we under-estimate the frequency of standing positions $\phi\approx0.5$. In this case 0.5 is considered the central point with 0 and 1 representing the left and right hoop respectively. We assumed that by introducing a difference in size between the two hoops that this distribution would shift towards the smaller hoop so as to balance their chances of success. We assumed that the difference in hoop size would cause particiapnts to stand slightly closer to the small hoop, in this case the shift in mean would be +0.05.

@@

The data from this experiment were analysed using a Bayesian beta regression. The recorded data for standing positions were transformed to be between 0 and 1, with 0 representing the larger hoop and 1 representing the smaller hoop. Therefore, the central point would be 0.5, and meaning anything above this value would demonstrate a shift in participant’s bhaviour away from the mid-point and towards the small hoop. The model included normalised hoop delta as a predictor to see how participants changed position with an increasing distance. This was also entered as a random effect by participant.

@@

This effect was the strongest for the closest separation which reflects the larger amount of variation in standing position with distance in the Two-throw condition (Figure \@ref(fig:TwoThrowPlts). In general, when in the One-throw condition, participants stood further from the centre (mean of `r round(mu_close[1], digits = 3)`, 95% HPDI of |`r round(lower_close[1], digits = 3)` , `r round(upper_close[1], digits = 3)`|) than in the the Two-throw condition (mean of `r round(mu_close[1], digits = 3)`, 95% HPDI of |`r round(lower_close[1], digits = 3)` , `r round(upper_close[1], digits = 3)`|) with P(One-throw > Two_throw|data) = `r paste(P_close$above0*100, "%", sep = "")`. However, as can be seen in \@ref(fig:TwoThrowPlts), the difference is generally small which and consitent across all distances. This means that when participants were given the opportunity to throw to both hoops (i.e., in the Two throw condition), they were still suboptimal in their performance.

@@

The priors in for this model were set so as to reduce the likelihood of extreme values (i.e., close to 0% or 100%). Until this experiment, we had not run a version of this task in which the different targets could have different value associtated with them.

As the prior predictive checks demonstrated that the model was able to retrieve the prior, and the priors were suitably flat, the model was conditioned on the raw data from the experiment. The model looked as followswith the prior predictive checks having the additional "sample_prior" argument set to "_only_":

@@

Probability matching may be due in part to a misunderstanding of probability [@REIMERS201811], but it has also been argued to reﬂect a reasonable tendency to seek out and exploit patterns in sequential events [@gaissmaier2008smart; @wolford2004searching]. @Gao2015 argue that when a person selects the more likely option on every trial (a strategy known as _Maximising_), they are accepting a loss for a minority of trials (20% in this example). As demonstrated in @KahnemanChoicesValuseFrames, people tend to ﬁnd certain loss very aversive, so a strategy that includes a certain loss may be disregarded as a “solution” to the problem at hand [@Goodnow1955]. The only way to avoid this loss is to make an attempt to detect and exploit any potential pattern. In the context of deciding where to ﬁxate between two potential targets locations, in both @morvan2012human and @clarke2015failure, the probability for each location to contain the target target was 50%. Given the philosophical and experimental arguments described above, participants may react to this choice between equally likely options with idiosyncratic pattern-seeking behavior [@gaissmaier2008smart; @yellott1969probability; @wolford2004searching]. In @clarke2015failure, this may have interfered with participants using information about their own ability to make decisions. Instead, they may have been attempting to ﬁgure out patterns in the sequence of targets to make better guesses about which one was likely to be the target on each trial. Adding in a bias for one option to be more likely may cater to people’s tendency to seek out, and attempt to exploit, patterns. It also breaks the Buridan’s Ass deadlock and makes the decision about which of two goals to attempt easier. In this Experiment, unequal probability was introduced into the paradigm used by @clarke2015failure in order to investigate 1) whether participants would make use of probability information in deciding where to ﬁxate, and 2) whether probability manipulations might facilitate the decision about whether to ﬁxate between two potential target locations or to look directly at one of them. If so, it would suggest that at least part of the reason for the poor decisions observed previously was the (unnatural) balance between the two potentialt argets. Additionally,this experiment will provide a useful replication of both the ﬁxations decision results [@morvan2012human; @clarke2015failure] and the probability matching tendency [@Vulkan2000ProbMatching] in a new context.

@@

For this experiment, data from ($N$ = 12) @clarke2015failure, which acted as a substitute for the Symmetric condition, and an unpublished pilot study ($N$ = 11), which followed the same rules as the Bias condition in the main experiment, were resampled in order to invesigate how often participants would fixate one of the side boxes given it was more likely to contain the target for the Bias condition. For the Symmetric condition, we followed the same rule as the in the paper and classified the most likely box as the side box that participants fixated most often (Figure \@ref(fig:ProbOGData). The main interest for our purposes was the proportion of time participants fixated the most likely box.

@@

The data were coded so that a fixation was classified as either being to the Most likely box or not. We then resampled (with replacement) these data by selecting a random $N$ participants from each condition (ranging from 2 to 20) then sampling 300 trials from these participants from the different datasets. This was done 5000 times to estimate the expected difference between the Symmetric and Bias conditions in terms of proportion of fixations to the Most likely side and the associated certainty around these values.

As can be seen in the figure below, the uncertainty surrounding the estimate for the difference between the groups appears to plateau somewhere around 15 participants. The shaded region represents a 95% Highest Density Interval (HDI) for the disribution of differences simulated through resampling. As such, the sample size of 18 in the main experiment appears to be ample in order to detect this difference and increasing the sample size above this value does not add more to the certainty, as can be seen in Figure \@ref(fig:ProbPowerPlots).

@@

```{r ProbPowerPlots, include = T, echo = F, fig.asp=.45, fig.cap="_Plot to show how the 95% HDI of the estimated difference in fixations to the most likely box changed with an increase in participants_"}
# maybe just the hdi stuff?
# do I want to rerun the sim with dist type added in?
plt_resampling <- df_sample %>% 
  spread(Condition, 
         Most) %>% 
  mutate(diff = Bias - Symmetric) %>% 
  group_by(n_sub) %>% 
  summarise(mu = mean(diff),
            lower = hdi(diff)[1],
            upper = hdi(diff)[2]) %>% 
  ggplot(aes(n_sub, mu)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lower,
                  ymax = upper),
              alpha = .3) +
  theme_bw() + 
  scale_x_continuous("Sample Size") + 
  scale_y_continuous("Differenc in fixations to more \n likely side (Bias - Symmetric)")

@@

### Participants
16 Particiapnts (two Male) were recruited from the University of Aberdeen community with an average age of 22.75 (between 19 and 29). Each participant was reimbursed £10 for their time.

@@

The experiment took place in a darkened room on a desktop computer. An Eyelink 1000 (version 4.594)(SR Research ltd, Mississauga, Ontario, Canada) was used to record eye position at 1000Hz. In each session, a 5 point calibration was carried out with additional calibration and validation sequences prior to each block, and if the participants had broken fixation ten times cummulatively or for five trials in a row. The stimuli were displayed on a CRT monitor (resolution 1920 $X$ 1080 pixels) using Matlab 7.9.0 (R2009b) with Psychtoolbox [@brainard1997psychophysics; @pelli1997videotoolbox] and EyelinkToolbox functions [@cornelissen2002eyelink]. A chin rest with forehead bar was used to ensure participants maintained a viewing distance of  $\approx$ 47cm.

In the first session, participants were instructed to remain ﬁxated at the center and identify a letterthatwouldappearinoneoftwoboxes. Atthestartofeachtrial,participantsﬁxatedacentral black cross on a grayscale background and pressed the spacebar. The cross was ﬂanked by two boxes which were a lighter shade of grey and occupied 1◦ of visual angle. After a stable ﬁxation had been maintained for 700ms, the target would appear in one of the two boxes for 500ms. The target was a white letter, 0.4&deg; of visual angle, drawn using the Sloan font as these letters are generally of equal recognisability at different viewing angles[@sloan1952comparison]. Ten letters were used and one was selected randomly on each trial to be the target letter. Participants were then presented with a screen prompting them to report which letter was presented by clicking on the corresponding character. An illustration of each trial in Session 1 can be seen in Figure \@ref(fig:ProbSession1).

@@

The optimal strategy is to ﬁxate whichever of the three boxes maximises expected accuracy. We refer to a participant’s expected accuracy if they had followed the optimal strategy, as “optimal accuracy”.  Using the psychometric curves ﬁt to each participant’s Session 1 data, we can calculate how likely a given person is to detect the target at various distances. This was then used to predict how accurate that same person would be if they had ﬁxated one of the side boxes, or the centre box, by using their average accuracy for a given distance from the left and right box ($Bl$ and $Br$ respectively) in Session 1. The ﬁxation choice that gave the greatest expected accuracy was then selectedastheoptimaldecisionfortherespectiveboxseparation. Asprobabilitywasalsoafactor in this experiment, this had to be accounted for by multiplying the chance of detecting the target in either box by the probability that the box would contain the target ($Pl$ and $Pr$). The formula for expected accuracy is therefore: $(Bl \times Pl) + (Br \times Pr)$. Expected accuracy given optimal ﬁxations differs between the Symmetric and Biased conditions. For example, in the Symmetric condition, participants could expect a 55% success rate if they ﬁxated the optimal location when the targets were far apart. This value comes from assuming they would be $\approx$ 100% accurate for the ﬁxated box, and at chance level for the non-ﬁxated box (10%). This gives us $(1 \times 0.5) + (0.1 \times 0.5) = 55%$. To get the lower limit for the Bias condition, we simply change the chance for each box to contain the target (additionally, we assume that the participant would ﬁxate the most likely box). This gives us $(1 \times 0.8) + (0.1 \times 0.2) = 82%$. This same formula can be used to calculate expected accuracy had the participant ﬁxated the central box. In this case, $Bl = Br$, and so would simply be put into the formula using the appropriate values for $P$, as demonstrated above.

@@

The choices made by participants on where to look are summarized in Figure \@ref(fig:ProbProp). The optimal strategy in this situation is to ﬁxate the central box when the two sides boxes are near (i.e., when their distance from centre, $\Delta$, is closer than each participant’s switchpoint). When they are far (i.e. delta is greater than the switch point), a participant behaving optimally should ﬁxate either of the side boxes when in the symmetric condition, and the most-likely box when in the biased condition. The ﬁrst panel of Figure \@ref(fig:ProbProp) clearly shows that our participants did not follow this strategy.  However, while adding a bias to the location of the target did not help participants to behave optimally, we can see that it did have an effect on their behavior, as they were much less likely to ﬁxate the central box. Furthermore, when ﬁxating one of the two side squares, they ﬁxated the square that was most likely to contain the target (Figure \@ref(fig:ProbProp), central and right panel) almost all the time

>>>>

These supplementary materials present full details of all R packages used in the analysis; a detailed explanation of our simulation-based power analysis; and Bayesian models of the results discussed in the main manuscript. In addition, we also present details a further, related experiment using eye-tracking and a probability matching paradigm.

@@

The fit with the empirical data is reasonable, although we under-estimate the frequency of standing positions $\phi\approx0.5$. In this case 0.5 is considered the central point with 0 and 1 representing the left and right hoop respectively. We assumed that by introducing a difference in size between the two hoops that this distribution would shift towards the smaller hoop so as to balance their chances of success. We assumed that the difference in hoop size would cause participants to stand slightly closer to the small hoop, in this case the shift in mean would be +0.05.

@@

The data from this experiment were analysed using a Bayesian beta regression. The recorded data for standing positions were transformed to be between 0 and 1, with 0 representing the larger hoop and 1 representing the smaller hoop. Therefore, the central point would be 0.5, and meaning anything above this value would demonstrate a shift in participant’s behavior away from the mid-point and towards the small hoop. The model included normalised hoop delta as a predictor to see how participants changed position with an increasing distance. This was also entered as a random effect by participant.

@@

This effect was the strongest for the closest separation which reflects the larger amount of variation in standing position with distance in the Two-throw condition (Figure \@ref(fig:TwoThrowPlts). In general, when in the One-throw condition, participants stood further from the centre (mean of `r round(mu_close[1], digits = 3)`, 95% HPDI of |`r round(lower_close[1], digits = 3)` , `r round(upper_close[1], digits = 3)`|) than in the the Two-throw condition (mean of `r round(mu_close[1], digits = 3)`, 95% HPDI of |`r round(lower_close[1], digits = 3)` , `r round(upper_close[1], digits = 3)`|) with P(One-throw > Two_throw|data) = `r paste(P_close$above0*100, "%", sep = "")`. However, as can be seen in \@ref(fig:TwoThrowPlts), the difference is generally small which and consistent across all distances. This means that when participants were given the opportunity to throw to both hoops (i.e., in the Two throw condition), they were still sub-optimal in their performance.

@@

The priors in for this model were set so as to reduce the likelihood of extreme values (i.e., close to 0% or 100%). Until this experiment, we had not run a version of this task in which the different targets could have different value associated with them.

As the prior predictive checks demonstrated that the model was able to retrieve the prior, and the priors were suitably flat, the model was conditioned on the raw data from the experiment. The model looked as follows with the prior predictive checks having the additional "sample_prior" argument set to "_only_":

@@

Probability matching may be due in part to a misunderstanding of probability [@REIMERS201811], but it has also been argued to reﬂect a reasonable tendency to seek out and exploit patterns in sequential events [@gaissmaier2008smart; @wolford2004searching]. @Gao2015 argue that when a person selects the more likely option on every trial (a strategy known as _Maximising_), they are accepting a loss for a minority of trials (20% in this example). As demonstrated in @KahnemanChoicesValuseFrames, people tend to ﬁnd certain loss very aversive, so a strategy that includes a certain loss may be disregarded as a “solution” to the problem at hand [@Goodnow1955]. The only way to avoid this loss is to make an attempt to detect and exploit any potential pattern. In the context of deciding where to ﬁxate between two potential targets locations, in both @morvan2012human and @clarke2015failure, the probability for each location to contain the target target was 50%. Given the philosophical and experimental arguments described above, participants may react to this choice between equally likely options with idiosyncratic pattern-seeking behavior [@gaissmaier2008smart; @yellott1969probability; @wolford2004searching]. In @clarke2015failure, this may have interfered with participants using information about their own ability to make decisions. Instead, they may have been attempting to ﬁgure out patterns in the sequence of targets to make better guesses about which one was likely to be the target on each trial. Adding in a bias for one option to be more likely may cater to people’s tendency to seek out, and attempt to exploit, patterns. It also breaks the Buridan’s Ass deadlock and makes the decision about which of two goals to attempt easier. In this Experiment, unequal probability was introduced into the paradigm used by @clarke2015failure in order to investigate 1) whether participants would make use of probability information in deciding where to ﬁxate, and 2) whether probability manipulations might facilitate the decision about whether to ﬁxate between two potential target locations or to look directly at one of them. If so, it would suggest that at least part of the reason for the poor decisions observed previously was the (unnatural) balance between the two potential targets. Additionally,this experiment will provide a useful replication of both the ﬁxations decision results [@morvan2012human; @clarke2015failure] and the probability matching tendency [@Vulkan2000ProbMatching] in a new context.

@@

For this experiment, data from ($N$ = 12) @clarke2015failure, which acted as a substitute for the Symmetric condition, and an unpublished pilot study ($N$ = 11), which followed the same rules as the Bias condition in the main experiment, were resampled in order to investigate how often participants would fixate one of the side boxes given it was more likely to contain the target for the Bias condition. For the Symmetric condition, we followed the same rule as the in the paper and classified the most likely box as the side box that participants fixated most often (Figure \@ref(fig:ProbOGData). The main interest for our purposes was the proportion of time participants fixated the most likely box.

@@

The data were coded so that a fixation was classified as either being to the Most likely box or not. We then resampled (with replacement) these data by selecting a random $N$ participants from each condition (ranging from 2 to 20) then sampling 300 trials from these participants from the different data sets. This was done 5000 times to estimate the expected difference between the Symmetric and Bias conditions in terms of proportion of fixations to the Most likely side and the associated certainty around these values.

As can be seen in the figure below, the uncertainty surrounding the estimate for the difference between the groups appears to plateau somewhere around 15 participants. The shaded region represents a 95% Highest Density Interval (HDI) for the distribution of differences simulated through resampling. As such, the sample size of 18 in the main experiment appears to be ample in order to detect this difference and increasing the sample size above this value does not add more to the certainty, as can be seen in Figure \@ref(fig:ProbPowerPlots).

@@

```{r ProbPowerPlots, include = T, echo = F, fig.asp=.45, fig.cap="_Plot to show how the 95% HDI of the estimated difference in fixations to the most likely box changed with an increase in participants_"}
# maybe just the hdi stuff?
# do I want to rerun the sim with dist type added in?
plt_resampling <- df_sample %>% 
  spread(Condition, 
         Most) %>% 
  mutate(diff = Bias - Symmetric) %>% 
  group_by(n_sub) %>% 
  summarise(mu = mean(diff),
            lower = hdi(diff)[1],
            upper = hdi(diff)[2]) %>% 
  ggplot(aes(n_sub, mu)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lower,
                  ymax = upper),
              alpha = .3) +
  theme_bw() + 
  scale_x_continuous("Sample Size") + 
  scale_y_continuous("Difference in fixations to more \n likely side (Bias - Symmetric)")

@@

### Participants
16 Participants (two Male) were recruited from the University of Aberdeen community with an average age of 22.75 (between 19 and 29). Each participant was reimbursed £10 for their time.

@@

The experiment took place in a darkened room on a desktop computer. An Eyelink 1000 (version 4.594)(SR Research ltd, Mississauga, Ontario, Canada) was used to record eye position at 1000Hz. In each session, a 5 point calibration was carried out with additional calibration and validation sequences prior to each block, and if the participants had broken fixation ten times cumulatively or for five trials in a row. The stimuli were displayed on a CRT monitor (resolution 1920 $X$ 1080 pixels) using Matlab 7.9.0 (R2009b) with Psychtoolbox [@brainard1997psychophysics; @pelli1997videotoolbox] and EyelinkToolbox functions [@cornelissen2002eyelink]. A chin rest with forehead bar was used to ensure participants maintained a viewing distance of  $\approx$ 47cm.

In the first session, participants were instructed to remain ﬁxated at the center and identify a letter that would appear in one of two boxes. At the start of each trial,participantsﬁxatedacentral black cross on a grayscale background and pressed the spacebar. The cross was ﬂanked by two boxes which were a lighter shade of grey and occupied 1◦ of visual angle. After a stable ﬁxation had been maintained for 700ms, the target would appear in one of the two boxes for 500ms. The target was a white letter, 0.4&deg; of visual angle, drawn using the Sloan font as these letters are generally of equal recognisability at different viewing angles[@sloan1952comparison]. Ten letters were used and one was selected randomly on each trial to be the target letter. Participants were then presented with a screen prompting them to report which letter was presented by clicking on the corresponding character. An illustration of each trial in Session 1 can be seen in Figure \@ref(fig:ProbSession1).

@@

The optimal strategy is to ﬁxate whichever of the three boxes maximises expected accuracy. We refer to a participant’s expected accuracy if they had followed the optimal strategy, as “optimal accuracy”.  Using the psychometric curves ﬁt to each participant’s Session 1 data, we can calculate how likely a given person is to detect the target at various distances. This was then used to predict how accurate that same person would be if they had ﬁxated one of the side boxes, or the centre box, by using their average accuracy for a given distance from the left and right box ($Bl$ and $Br$ respectively) in Session 1. The ﬁxation choice that gave the greatest expected accuracy was then selected as the optimal decision for the respective box separation. As probability was also a factor in this experiment, this had to be accounted for by multiplying the chance of detecting the target in either box by the probability that the box would contain the target ($Pl$ and $Pr$). The formula for expected accuracy is therefore: $(Bl \times Pl) + (Br \times Pr)$. Expected accuracy given optimal ﬁxations differs between the Symmetric and Biased conditions. For example, in the Symmetric condition, participants could expect a 55% success rate if they ﬁxated the optimal location when the targets were far apart. This value comes from assuming they would be $\approx$ 100% accurate for the ﬁxated box, and at chance level for the non-ﬁxated box (10%). This gives us $(1 \times 0.5) + (0.1 \times 0.5) = 55%$. To get the lower limit for the Bias condition, we simply change the chance for each box to contain the target (additionally, we assume that the participant would ﬁxate the most likely box). This gives us $(1 \times 0.8) + (0.1 \times 0.2) = 82%$. This same formula can be used to calculate expected accuracy had the participant ﬁxated the central box. In this case, $Bl = Br$, and so would simply be put into the formula using the appropriate values for $P$, as demonstrated above.

@@

The choices made by participants on where to look are summarized in Figure \@ref(fig:ProbProp). The optimal strategy in this situation is to ﬁxate the central box when the two sides boxes are near (i.e., when their distance from centre, $\Delta$, is closer than each participant’s switch point). When they are far (i.e. delta is greater than the switch point), a participant behaving optimally should ﬁxate either of the side boxes when in the symmetric condition, and the most-likely box when in the biased condition. The ﬁrst panel of Figure \@ref(fig:ProbProp) clearly shows that our participants did not follow this strategy.  However, while adding a bias to the location of the target did not help participants to behave optimally, we can see that it did have an effect on their behavior, as they were much less likely to ﬁxate the central box. Furthermore, when ﬁxating one of the two side squares, they ﬁxated the square that was most likely to contain the target (Figure \@ref(fig:ProbProp), central and right panel) almost all the time
